Four decades of water quality change in the upper San Francisco Estuary

### 2017-05-02 sftrends_manu.tex
cover letter added, manuscript and supp info separated

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary\footnote{Version: Tue May 2 08:37:44 2017 -0500, \href{https://github.com/fawda123/sftrends_manu/commit/2a1fcbeba412e8ea16c8fea91f775c46617d667d}{2a1fcbeba412e8ea16c8fea91f775c46617d667d}\newline The views expressed in this paper are those of the authors and do not necessarily reflect the views or policies of the U.S. Environmental Protection Agency.}}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing). P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section. Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (\cref{fig:trndcomp2}) and decreases were observed for 1996-2013 (\cref{fig:trndcomp3}).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects. As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in \cref{tab:trndsann}) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.  Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (\cref{fig:trndcomp2,fig:trndcomp3,tab:trndsbef,tab:trndsaft}).  The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

@@

Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (\cref{fig:stock}).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different annual groupings. See \cref{fig:trndcomp2,fig:trndcomp3} for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

@@

\caption{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.  Nitrite/nitrate trends are on the left and ammonium trends are on the right.  Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (\cref{fig:stock}).}\label{fig:p8trnds}
\end{figure}

@@

% trends in no23, nh at p8
%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg", insert.bottom = foot.val)%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8. Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, \cref{fig:stock}).  Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).  Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Increasing values are in bold-italics. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:p8chg}} 
\begin{center}
\begin{tabular}{lllcll}
\hline\hline
\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
\cline{2-3} \cline{5-6}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
\hline
{\bfseries Annual}&&&&&\tabularnewline
~~1976-2006&1.3&\textit{\textbf{2}}**&&0.2&\textit{\textbf{2.8}}**\tabularnewline
~~2007-2013&1.4&-1.9**&&0.1&-16.6**\tabularnewline
\hline
{\bfseries Seasonal, pre}&&&&&\tabularnewline
~~Spring&1.2&\textit{\textbf{1.6}}**&&0.2&\textit{\textbf{1.4}}**\tabularnewline
~~Summer&1&\textit{\textbf{2.4}}**&&0.1&\textit{\textbf{3.3}}**\tabularnewline
~~Fall&1.3&\textit{\textbf{2.2}}**&&0.2&\textit{\textbf{4.9}}**\tabularnewline
~~Winter&1.5&\textit{\textbf{2.1}}**&&0.7&\textit{\textbf{4.8}}**\tabularnewline
\hline
{\bfseries Seasonal, post}&&&&&\tabularnewline
~~Spring&1.3&-1.6**&&0.1&-16.2**\tabularnewline
~~Summer&0.9&-4.3**&&0.1&-15.7**\tabularnewline
~~Fall&1.5&-1.7**&&0.1&-19.3**\tabularnewline
~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
\hline
\end{tabular}\end{center}

\footnotesize *$p<0.05$; **$p<0.005$\end{table}

@@

% supplementary material
\begin{suppinfo}
\beginsupplement

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp2.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different seasonal groupings from 1976-1995. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See \cref{fig:trndcomp1} for annual comparisons.}
\label{fig:trndcomp2}   
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp3.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different seasonal groupings from 1996-2013. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF. See \cref{fig:trndcomp1} for annual comparisons.}
\label{fig:trndcomp3}   
\end{figure}

{\centering \includegraphics[width=\textwidth]{figs/stock-1}

\caption[Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County]{Nitrogen concentration measurements (mg L$^{-1}$) from the City of Stockton Wastewater Treatment Plant, San Joaquin County.  Wastewater discharge requirements were implemented in 2006 for nitrification/denitrification and tertiary filtration to convert ammonium to nitrate.}\label{fig:stock}
\end{figure}

% overall trends, years
%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Annual"), n.cgroup = c(2), rowname = stats,     label = "tab:trndsann", insert.bottom = foot.val)%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and annual aggregations.  Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period.\label{tab:trndsann}} 
\begin{center}
\begin{tabular}{lll}
\hline\hline
\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{2}{c}{\bfseries Annual}\tabularnewline
\cline{2-3}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{1976-1995}&\multicolumn{1}{c}{1996-2013}\tabularnewline
\hline
{\bfseries DIN}&&\tabularnewline
~~C10&1.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.4 \footnotesize{(-3.1)**}\tabularnewline
~~C3&0.3 \footnotesize{(\textit{\textbf{2.2}})**}&0.5 \footnotesize{(-0.1)**}\tabularnewline
~~D19&0.4 \footnotesize{(\textit{\textbf{0.2}})**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
~~D26&0.4 \footnotesize{(\textit{\textbf{0.4}})**}&0.5 \footnotesize{(-1.2)**}\tabularnewline
~~D28&0.4 \footnotesize{(\textit{\textbf{0.1}})**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
~~D4&0.3 \footnotesize{(\textit{\textbf{0.6}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
~~D6&0.4 \footnotesize{(\textit{\textbf{1.8}})**}&0.5 \footnotesize{(-0.3)**}\tabularnewline
~~D7&0.4 \footnotesize{(\textit{\textbf{1.7}})**}&0.5 \footnotesize{(-0.7)**}\tabularnewline
~~MD10&0.4 \footnotesize{(-1.1)**}&0.3 \footnotesize{(-2.4)**}\tabularnewline
~~P8&1.3 \footnotesize{(\textit{\textbf{2.5}})**}&1.7 \footnotesize{(-2)**}\tabularnewline
\hline
{\bfseries NH$_{4}^{+}$}&&\tabularnewline
~~C10&0.1 \footnotesize{(-3.4)**}&0 \footnotesize{(-5.2)**}\tabularnewline
~~C3&0.2 \footnotesize{(\textit{\textbf{3.7}})**}&0.3 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
~~D19&0 \footnotesize{(\textit{\textbf{0.4}})**}&0 \footnotesize{(-1.7)**}\tabularnewline
~~D26&0.1 \footnotesize{(\textit{\textbf{2.2}})**}&0.1 \footnotesize{(-1.5)**}\tabularnewline
~~D28&0 \footnotesize{(-1.1)**}&0 \footnotesize{(-1.4)**}\tabularnewline
~~D4&0 \footnotesize{(\textit{\textbf{0.9}})**}&0.1 \footnotesize{(\textit{\textbf{0}})}\tabularnewline
~~D6&0.1 \footnotesize{(\textit{\textbf{2.4}})**}&0.1 \footnotesize{(-0.5)**}\tabularnewline
~~D7&0.1 \footnotesize{(\textit{\textbf{1.5}})**}&0.1 \footnotesize{(-1.2)**}\tabularnewline
~~MD10&0.1 \footnotesize{(-2.8)**}&0 \footnotesize{(-1.1)**}\tabularnewline
~~P8&0.2 \footnotesize{(\textit{\textbf{4.9}})**}&0.1 \footnotesize{(-10.3)**}\tabularnewline
\hline
{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&\tabularnewline
~~C10&1.2 \footnotesize{(\textit{\textbf{1.4}})**}&1.4 \footnotesize{(-3)**}\tabularnewline
~~C3&0.1 \footnotesize{(-0.1)**}&0.2 \footnotesize{(\textit{\textbf{0.7}})**}\tabularnewline
~~D19&0.4 \footnotesize{(-0.1)**}&0.4 \footnotesize{(-1.9)**}\tabularnewline
~~D26&0.3 \footnotesize{(\textit{\textbf{0}})}&0.4 \footnotesize{(-1.1)**}\tabularnewline
~~D28&0.4 \footnotesize{(-0.2)**}&0.4 \footnotesize{(-3.1)**}\tabularnewline
~~D4&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(-0.4)**}\tabularnewline
~~D6&0.3 \footnotesize{(\textit{\textbf{1.3}})**}&0.4 \footnotesize{(-0.3)**}\tabularnewline
~~D7&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.4 \footnotesize{(-0.7)**}\tabularnewline
~~MD10&0.4 \footnotesize{(-1)**}&0.3 \footnotesize{(-2.5)**}\tabularnewline
~~P8&1.2 \footnotesize{(\textit{\textbf{1.7}})**}&1.5 \footnotesize{(-0.6)**}\tabularnewline
\hline
\end{tabular}\end{center}

\footnotesize *$p<0.05$; **$p<0.005$\end{table}

% overall trends, seasons, first annual periodyears
%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1976-1995"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsbef", insert.bottom = foot.val)%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1976-1995. Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsbef}} 
\begin{center}
\begin{tabular}{lllll}
\hline\hline
\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1976-1995}\tabularnewline
\cline{2-5}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
\hline
{\bfseries DIN}&&&&\tabularnewline
~~C10&1.2 \footnotesize{(\textit{\textbf{1.1}})**}&1.2 \footnotesize{(\textit{\textbf{0.3}})}&1.3 \footnotesize{(\textit{\textbf{0.5}})**}&1.7 \footnotesize{(\textit{\textbf{1.2}})**}\tabularnewline
~~C3&0.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.3 \footnotesize{(\textit{\textbf{2.3}})**}&0.4 \footnotesize{(\textit{\textbf{2.4}})**}&0.4 \footnotesize{(\textit{\textbf{1.9}})**}\tabularnewline
~~D19&0.5 \footnotesize{(\textit{\textbf{0.3}})}&0.2 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{0.7}})**}&0.7 \footnotesize{(-0.2)}\tabularnewline
~~D26&0.4 \footnotesize{(\textit{\textbf{0.7}})**}&0.3 \footnotesize{(\textit{\textbf{0.4}})*}&0.4 \footnotesize{(\textit{\textbf{1}})**}&0.6 \footnotesize{(\textit{\textbf{0.3}})}\tabularnewline
~~D28&0.5 \footnotesize{(\textit{\textbf{0.8}})*}&0.2 \footnotesize{(\textit{\textbf{0.3}})}&0.3 \footnotesize{(\textit{\textbf{0.5}})*}&0.8 \footnotesize{(-0.3)}\tabularnewline
~~D4&0.4 \footnotesize{(\textit{\textbf{0.2}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.5 \footnotesize{(-0.5)}\tabularnewline
~~D6&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.3 \footnotesize{(\textit{\textbf{4.6}})**}&0.4 \footnotesize{(\textit{\textbf{1.4}})**}&0.5 \footnotesize{(-0.7)*}\tabularnewline
~~D7&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.2}})**}&0.4 \footnotesize{(\textit{\textbf{1.5}})**}&0.6 \footnotesize{(-2.4)**}\tabularnewline
~~MD10&0.6 \footnotesize{(-0.3)}&0.2 \footnotesize{(-3.6)**}&0.3 \footnotesize{(\textit{\textbf{0.8}})**}&1.3 \footnotesize{(-0.3)*}\tabularnewline
~~P8&1.3 \footnotesize{(\textit{\textbf{2.4}})**}&0.9 \footnotesize{(\textit{\textbf{2.4}})**}&1.3 \footnotesize{(\textit{\textbf{3.1}})**}&1.9 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
\hline
{\bfseries NH$_{4}^{+}$}&&&&\tabularnewline
~~C10&0.1 \footnotesize{(-2.3)**}&0 \footnotesize{(-6.8)**}&0.1 \footnotesize{(-7.1)**}&0.3 \footnotesize{(-1.5)**}\tabularnewline
~~C3&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.2 \footnotesize{(\textit{\textbf{4}})**}&0.3 \footnotesize{(\textit{\textbf{3.8}})**}&0.2 \footnotesize{(\textit{\textbf{2.9}})**}\tabularnewline
~~D19&0.1 \footnotesize{(\textit{\textbf{0.4}})*}&0 \footnotesize{(-1.7)**}&0 \footnotesize{(\textit{\textbf{1.2}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}\tabularnewline
~~D26&0.1 \footnotesize{(\textit{\textbf{1.4}})**}&0.1 \footnotesize{(\textit{\textbf{2.5}})**}&0.1 \footnotesize{(\textit{\textbf{3.1}})**}&0.1 \footnotesize{(\textit{\textbf{2.3}})**}\tabularnewline
~~D28&0.1 \footnotesize{(-0.5)}&0 \footnotesize{(-3.7)**}&0 \footnotesize{(-1.6)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
~~D4&0.1 \footnotesize{(\textit{\textbf{1.7}})**}&0 \footnotesize{(\textit{\textbf{1}})**}&0 \footnotesize{(-0.7)}&0.1 \footnotesize{(\textit{\textbf{2}})**}\tabularnewline
~~D6&0.1 \footnotesize{(\textit{\textbf{2.9}})**}&0.1 \footnotesize{(\textit{\textbf{2.8}})**}&0.1 \footnotesize{(-0.1)}&0.1 \footnotesize{(\textit{\textbf{2.1}})**}\tabularnewline
~~D7&0.1 \footnotesize{(\textit{\textbf{3.3}})**}&0 \footnotesize{(\textit{\textbf{2}})**}&0.1 \footnotesize{(-2.8)**}&0.1 \footnotesize{(\textit{\textbf{1.7}})**}\tabularnewline
~~MD10&0.1 \footnotesize{(-1.8)**}&0 \footnotesize{(-6.5)**}&0 \footnotesize{(-3.3)**}&0.2 \footnotesize{(\textit{\textbf{0.4}})}\tabularnewline
~~P8&0.2 \footnotesize{(\textit{\textbf{3.9}})**}&0.1 \footnotesize{(\textit{\textbf{1.8}})**}&0.2 \footnotesize{(\textit{\textbf{7}})**}&0.6 \footnotesize{(\textit{\textbf{7}})**}\tabularnewline
\hline
{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&&&&\tabularnewline
~~C10&1.1 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(\textit{\textbf{0.6}})**}&1.2 \footnotesize{(\textit{\textbf{1.3}})**}&1.5 \footnotesize{(\textit{\textbf{1.8}})**}\tabularnewline
~~C3&0.2 \footnotesize{(\textit{\textbf{0.7}})**}&0.1 \footnotesize{(-1)**}&0.1 \footnotesize{(-0.3)}&0.2 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
~~D19&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.2 \footnotesize{(-0.3)}&0.3 \footnotesize{(\textit{\textbf{0.3}})}&0.6 \footnotesize{(-0.9)*}\tabularnewline
~~D26&0.4 \footnotesize{(\textit{\textbf{0.6}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.3}})*}&0.5 \footnotesize{(-0.3)}\tabularnewline
~~D28&0.5 \footnotesize{(\textit{\textbf{0.7}})*}&0.2 \footnotesize{(-0.1)}&0.3 \footnotesize{(\textit{\textbf{0.2}})}&0.7 \footnotesize{(-1)**}\tabularnewline
~~D4&0.3 \footnotesize{(\textit{\textbf{0.1}})}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.3 \footnotesize{(\textit{\textbf{1.1}})**}&0.4 \footnotesize{(-0.8)*}\tabularnewline
~~D6&0.4 \footnotesize{(-0.2)}&0.3 \footnotesize{(\textit{\textbf{4.1}})**}&0.3 \footnotesize{(\textit{\textbf{1.4}})**}&0.4 \footnotesize{(-1)**}\tabularnewline
~~D7&0.4 \footnotesize{(-1)*}&0.3 \footnotesize{(\textit{\textbf{3.4}})**}&0.4 \footnotesize{(\textit{\textbf{0.4}})}&0.4 \footnotesize{(-3.6)**}\tabularnewline
~~MD10&0.5 \footnotesize{(-0.2)}&0.2 \footnotesize{(-3.6)**}&0.2 \footnotesize{(\textit{\textbf{1.5}})**}&1.2 \footnotesize{(-0.5)*}\tabularnewline
~~P8&1.2 \footnotesize{(\textit{\textbf{2}})**}&0.9 \footnotesize{(\textit{\textbf{2.3}})**}&1.1 \footnotesize{(\textit{\textbf{2}})**}&1.4 \footnotesize{(\textit{\textbf{1}})**}\tabularnewline
\hline
\end{tabular}\end{center}

\footnotesize *$p<0.05$; **$p<0.005$\end{table}

% overall trends, seasons, first annual periodyears
%latex.default(tab[, -c(1, 2)], file = "", rowlabel = "Analyte/Station",     caption = cap.val, caption.loc = "top", rgroup = vars, n.rgroup = rep(nrow(tab)/3,         3), cgroup = c("Seasonal, 1996-2013"), n.cgroup = c(4),     rowname = stats, label = "tab:trndsaft", insert.bottom = foot.val)%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in nitrogen analytes for all stations and seasonal aggregations from 1996-2013. Summaries are  medians (mg L$^{-1}$) and percent change per year in parentheses (increasing in bold-italic). Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:trndsaft}} 
\begin{center}
\begin{tabular}{lllll}
\hline\hline
\multicolumn{1}{l}{\bfseries Analyte/Station}&\multicolumn{4}{c}{\bfseries Seasonal, 1996-2013}\tabularnewline
\cline{2-5}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Spring}&\multicolumn{1}{c}{Summer}&\multicolumn{1}{c}{Fall}&\multicolumn{1}{c}{Winter}\tabularnewline
\hline
{\bfseries DIN}&&&&\tabularnewline
~~C10&1.1 \footnotesize{(-4.1)**}&1.3 \footnotesize{(-3.1)**}&1.6 \footnotesize{(-2)**}&1.7 \footnotesize{(-3.4)**}\tabularnewline

>>>>

\title[Water quality change in the upper SFE]{Four decades of water quality change in the upper San Francisco Estuary}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\linenumbers

@@

Application of seasonal Kendall tests to evaluate trends in observed data provided explicit information on the direction, magnitude, and statistical significance of changes between years. Trends estimated from the observed data for 1976-1995 and 1996-2013 varied considerably between sites and analytes (\cref{fig:trndcomp1}). Significant trends were observed from 1976-1995 for eight of ten sites for \ac{din} (seven increasing, one decreasing), eight sites for ammonium (six increasing, two decreasing), and six sites for nitrite/nitrate (five increasing, one decreasing).  Decreasing trends were more common for the observed data from 1996-2013.  Eight sites had significant trends for \ac{din} (four increasing, four decreasing), seven sites for ammonium (five increasing, two decreasing), and eight sites for nitrite/nitrate (four increasing, four decreasing). P8 had a relatively large decrease in ammonium ($-8.3$\% change per year) for the second annual period compared to all other sites, which is described in detail in the next section. Trends by season were similar such that increases were generally observed in all seasons from 1976-1995 (Figure S1) and decreases were observed for 1996-2013 (Figure S2).  Trends for the seasonal comparisons were noisier and significant changes were less common compared to the annual comparisons.

Relationships between flow and observed water quality are complex and can change significantly through space and time\cite{Hirsch10,Zhang16}. These principles have been demonstrated for monitoring data in the Delta region\cite{Jassby08,Novick15,Jabusch16}, suggesting that trend analyses using the observed time series are confounded by flow effects. As such, a comparison of flow-normalized results from \ac{wrtds} relative to observed data identified changes in the magnitude, significance, and direction of trends. For all sixty trend comparisons in \cref{fig:trndcomp1} (flow-normalized values in Table S1) regardless of site, nitrogen analyte, and time period, thirteen comparisons had trends that were insignificant with the observed data but significant with flow-normalized results, whereas only one trend changed to insignificant. This suggests that time series that include flow effects had sufficient noise to obscure or prevent identification of an actual trend of a water quality parameter. Further, changes in the magnitude of the estimated percent change per year were also apparent for the flow-normalized trends, such that fourteen comparisons showed an increase in magnitude (more negative or more positive) and twenty five had a decrease (less positive or less negative) compared to observed trends.  Eleven comparisons showed a trend reversal from positive to negative estimated change, nine sites went from no change to negative estimated change, and one site went from no change to a positive trend for the flow-normalized results. Differences by season in the observed relative to flow-normalized trends from \ac{wrtds} were also apparent (Figures S1 and S2 and Tables S2 and S3). The most notable changes were an overall decrease in the estimated trend for most sites in the summer and fall seasons for 1996-2013, including an increase in the number of statistically significant trends.

@@

Effluent concentrations measured from 2003 to 2009 from the Stockton \ac{wwtp} showed a gradual reduction in ammonium concentration relative to total \ac{din} (Figure S3).  Ammonium and nitrate concentrations were comparable prior to 2006, whereas nitrate was a majority of total nitrogen after the upgrade, with much smaller percentages from ammonium and nitrite. As expected, flow-normalized nitrogen trends at P8 shifted in response to upstream \ac{wwtp} upgrades (\cref{fig:p8trnds}a), with ammonium showing an increase from 1976 followed by a large reduction in the 2000s.  Interestingly, nitrite/nitrate concentrations also showed a similar but less dramatic decrease despite an increase in the \ac{wwtp} effluent concentrations following the upgrade.  Percent changes from seasonal Kendall tests on flow-normalized results showed that both nitrogen species increased prior to \ac{wwtp} upgrades ($2$\% per year for nitrite/nitrate, $2.8$\% for ammonium), followed by decreases after upgrades ($-1.9$\% for nitrite/nitrate, $-16.6$\% for ammonium, \cref{tab:p8chg}).  Seasonally, increases prior to upgrades were highest in the summer for nitrite/nitrate ($2.4$\%) and in the fall for ammonium ($4.9$\%). Similarly, seasonal reductions post-upgrade were largest in the summer for nitrite/nitrate ($-4.3$\%) and largest for ammonium in the winter ($-26.7$\%).

@@

\begin{figure}
\centering
\includegraphics[width=1\textwidth,page=1]{figs/trndcomp1.pdf}
\caption{Results from seasonal Kendall tests on observed data (triangles) and flow-normalized predictions (circles) from \ac{wrtds} for nitrogen analytes. Results are shown as the percent change per year as the estimated Theil-Sen slope divided by the median for a given aggregation period (significance evaluated at $\alpha = 0.05$, based on $\tau$). Trends are shown separately for different annual groupings. See Figures S1 and S2 for seasonal groupings.}
\label{fig:trndcomp1}   
\end{figure}

@@

\caption[Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}]{Nitrogen trends at P8 as (a, top) observed, (a, bottom) predicted and flow-normalized estimates from \ac{wrtds}, and (b) relationships with flow over time from \ac{wrtds}.  Nitrite/nitrate trends are on the left and ammonium trends are on the right.  Wastewater treatment plant upgrades at the City of Stockton (San Joaquin County) were completed in 2006 (Figure S3).}\label{fig:p8trnds}
\end{figure}

@@

% trends in no23, nh at p8
%latex.default(totab, file = "", rowlabel = "Period", caption = cap.val,     caption.loc = "top", rgroup = segs, n.rgroup = c(2, rep(4,         2)), cgroup = vars, n.cgroup = c(2, 2), rowname = cats,     label = "tab:p8chg", insert.bottom = foot.val)%
\begin{table}[!tbp]
\caption{Summaries of flow-normalized trends in nitrite/nitrate and ammonium (mg L$^{-1}$) concentrations before and after \ac{wwtp} upgrades upstream of station P8. Upgrades were completed in 2006 at the City of Stockton \ac{wwtp} (San Joaquin County, Figure S3).  Summaries are  medians and percent change per year in parentheses (increasing in bold-italic).  Changes and significance estimates are based on seasonal Kendall tests of flow-normalized results within each time period. Increasing values are in bold-italics. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.\label{tab:p8chg}} 
\begin{center}
\begin{tabular}{lllcll}
\hline\hline
\multicolumn{1}{l}{\bfseries Period}&\multicolumn{2}{c}{\bfseries NO$_{2}^{-}$/NO$_{3}^{2-}$}&\multicolumn{1}{c}{\bfseries }&\multicolumn{2}{c}{\bfseries NH$_{4}^{+}$}\tabularnewline
\cline{2-3} \cline{5-6}
\multicolumn{1}{l}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}&\multicolumn{1}{c}{}&\multicolumn{1}{c}{Median}&\multicolumn{1}{c}{\% change}\tabularnewline
\hline
{\bfseries Annual}&&&&&\tabularnewline
~~1976-2006&1.3&\textit{\textbf{2}}**&&0.2&\textit{\textbf{2.8}}**\tabularnewline
~~2007-2013&1.4&-1.9**&&0.1&-16.6**\tabularnewline
\hline
{\bfseries Seasonal, pre}&&&&&\tabularnewline
~~Spring&1.2&\textit{\textbf{1.6}}**&&0.2&\textit{\textbf{1.4}}**\tabularnewline
~~Summer&1&\textit{\textbf{2.4}}**&&0.1&\textit{\textbf{3.3}}**\tabularnewline
~~Fall&1.3&\textit{\textbf{2.2}}**&&0.2&\textit{\textbf{4.9}}**\tabularnewline
~~Winter&1.5&\textit{\textbf{2.1}}**&&0.7&\textit{\textbf{4.8}}**\tabularnewline
\hline
{\bfseries Seasonal, post}&&&&&\tabularnewline
~~Spring&1.3&-1.6**&&0.1&-16.2**\tabularnewline
~~Summer&0.9&-4.3**&&0.1&-15.7**\tabularnewline
~~Fall&1.5&-1.7**&&0.1&-19.3**\tabularnewline
~~Winter&2.2&-0.8**&&0.2&-26.7**\tabularnewline
\hline
\end{tabular}\end{center}
\footnotesize *$p<0.05$; **$p<0.005$\end{table}

@@
